---
title: "Random Networks"
author: "Jake L. Weissman"
date: "May 21, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Today we are going to generate some random networks. Colloquially, when people 
say "random" they often mean "uniformly distributed" rather than "stochastic". 
Importantly, there are many different ways to stochastically generate a network,
and we will discuss a few day.

The point of this lesson is to get you think about the processes generating your
own networks, and how you might characterize them (and give you a couple places
to look when starting with these analyses). If you want a rigorous 
mathematical treatment of random networks I recommend any of the textbooks/courses
listed in the extended resources section on the wiki (most textbooks dedicate a
significant amount of space to these topics). **Importantly, a typical graduate-level
network science course would spend several weeks at a minimum on the topics I introduce here.**

First let's load in the packages we will be working with. This lesson will mostly
use `igraph`, though we again use the example gut network provided with `SpiecEasi`.
I'll also briefly introduce the `randnet` package at the end of the lesson for
fitting stochastic block models.

```{r Load Packages}
#Make sure installed
if("igraph" %in% rownames(installed.packages()) == FALSE){install.packages("igraph")}
if("SpiecEasi" %in% rownames(installed.packages()) == FALSE){devtools::install_github("zdk123/SpiecEasi")}
if("randnet" %in% rownames(installed.packages()) == FALSE){install.packages("randnet")}

#load
library(SpiecEasi)
library(igraph)
library(randnet)
```

## Erdos-Reyni Networks

The Erdos-Reyni random network model is what people usually mean when they say
"random network". This type of network can be generated in two ways:  
1. Setting a number of edges `L` and randomly placing them between nodes  
2. Setting a probability `p` that any two nodes are connected by an edge

We will use approach 2, which is a little easier to work with mathematically and is
more widely used (though #1 is the original definition). 

Let's give our network 100 nodes, and a probability of 0.1 that any given pair of 
nodes is connected (each node should have around one connection, a bit less).

```{r ER}
n <- 100
p <- 0.1
er.net <- erdos.renyi.game(n, p, type = "gnp")
plot(er.net,
     vertex.label=NA,
     vertex.size=1)
```

Not particularly exciting as far as networks go (not much interesting structure in
terms of hubs, clusters). Even so, there are some interesting properties of this
type of network. For example, there is a "phase transition" around p=0.01 where
the network goes from mostly disconnected to mostly connected in a single component
(look up percolation theory if you are interested in learning more).

```{r percolation}
n <- 100
p <- 0.01
er.net <- erdos.renyi.game(n, p, type = "gnp")
plot(er.net,
     vertex.label=NA,
     vertex.size=1)

n <- 100
p <- 0.02
er.net <- erdos.renyi.game(n, p, type = "gnp")
plot(er.net,
     vertex.label=NA,
     vertex.size=1)
```

One way people often analyze network structure is by looking at the degree distribution
(the probability distribution of how likely a node is to have a specific degree).

```{r degree dist}
# Really big, really sparse Erdos-Reyni Network
n <- 10000
p <- 0.01
er.net <- erdos.renyi.game(n, p, type = "gnp")

degree_df <- data.frame(k = 1:length(degree.distribution(er.net)) - 1,
                        prob = degree.distribution(er.net))

plot(prob ~ k, data=degree_df)
```

For random network models, there is usually an expected degree distribution 
associated with the model. For example, for sparse Erdos-Reyni networks the
degree distribution is well approximated by a Poisson distribution with rate
parameter lambda equal to the mean degree of the network (which should equal p(n-1)).

```{r poisson}
k_mean <- p*(n-1)
theoretical_degree_df <- data.frame(k=1:150,
                                    prob=dpois(x=1:150,lambda=k_mean))
plot(prob ~ k, data=degree_df)
lines(prob ~ k, data=theoretical_degree_df)
```

This is one good way to check if your network resembles an Erdos-Reyni network or
has some immediately-apparent complicated structure.

## The Barabasi-Albert Model (Preferential Attachement)

Often, the networks we are interested in have "hubs" - specific nodes with very
high degree. You won't see these hubs form with the Erdos-Reyni model above, since
all nodes tend to have similar degree. Networks with hubs generally have "heavy-tailed"
degree distributions (I'll show an example in a moment).

One simple way to generate a random network with hubs is the Barabasi-Albert algorithm.
Under this algorithm you start with a seed network and iteratively add nodes. For
each added node, you connect it to other nodes in the network with a probability
proportional to their degree (see [the Barabasi textbook](http://networksciencebook.com/chapter/5#barabasi-model) for details). Thus
nodes with high degree accumulate new connections (a "rich-get-richer" type scenario),
leading to the formation of hubs.

Let's look at an example

```{r BA}
n <- 100
ba.net <- barabasi.game(n, directed = F)
plot(ba.net,
     vertex.label=NA,
     vertex.size=1)
```

What does the degree distribution of a network built under this algorithm look like?

```{r BADD}
n <- 10000
ba.net <- barabasi.game(n, directed = F)
degree_df <- data.frame(k = 1:length(degree.distribution(ba.net)) - 1,
                        prob = degree.distribution(ba.net))

k_mean <- mean(degree(ba.net))
theoretical_degree_df <- data.frame(k=1:150,
                                    prob=dpois(x=1:150,lambda=k_mean))
plot(prob ~ k, data=degree_df,log="xy")
lines(prob ~ k, data=theoretical_degree_df,log="xy")
```

The curved line is the theoretical fit of the poisson distribution 
(unlike with the E-R model, it is NOT a good fit here). 
Notice that we have log-log scaled our data, and get an approximately linear distribution. This corresponds to a "power-law", and network scientists would refer to this as a "scale-free" distribution
(meaning it has infinite variance, i.e., no "scale" of observation captures it; 
honestly, just start with the [wikipedia page](https://en.wikipedia.org/wiki/Scale-free_network)).

Fitting a power law may seem simple (it's just a line on a log-log plot), but you
should proceed with caution. Many have tried to fit power laws without being rigorous.
I'm not going to get into it here, but the [powRlaw](https://cran.r-project.org/web/packages/poweRlaw/index.html) has a nice [vignette](https://cran.r-project.org/web/packages/poweRlaw/vignettes/b_powerlaw_examples.pdf) if you 
are interested.

There are a LOT of variants of/extensions to the BA model (see the [Barabasi Textbook](http://networksciencebook.com/chapter/6#evolving-networks)). 
Feel free to read up on them if you want. Its worth thinking about what processes
are leading to the network you are studying (even if just as an intellectual excercise).

## Stochastic Block Models

Something we've talked about quite a bit so far in office hours is community-detection.
So how do you generate a random network with communities? 

The stochastic block model is basically the Erdos-Reyni model with communities.
You pre-specify some number of communities *K* and a *KxK* matrix whose entries
describe the probability of edges within (on the diagonal) or between (off the diagonal)
communities. Basically, its a bunch of Erdos-Reyni networks that you then connect
at random.

```{r SBM}
pm <- cbind( c(.3, .01), c(.01, .5) )
sbm.net <- sample_sbm(n=100, pref.matrix=pm, block.sizes=c(30,70))
plot(sbm.net,
     vertex.label=NA,
     vertex.size=1)
```

What do their degree distributions look like? Well, this one is bimodal (one peak
for each cluster - each peak roughly Poisson-distributed) - why do you think our
theoretical prediction is a little off for the lower peak?

```{r SBM DD}
pm <- cbind( c(.3, .01), c(.01, .5) )
sbm.net <- sample_sbm(n=10000, pref.matrix=pm, block.sizes=c(3000,7000))
degree_df <- data.frame(k = 1:length(degree.distribution(sbm.net)) - 1,
                        prob = degree.distribution(sbm.net))

k_mean1 <- 0.3*(3000-1)
theoretical_degree_df1 <- data.frame(k=1:5000,
                                    prob=dpois(x=1:5000,lambda=k_mean1))
k_mean2 <- 0.5*(7000-1)
theoretical_degree_df2 <- data.frame(k=1:5000,
                                    prob=dpois(x=1:5000,lambda=k_mean2))

plot(prob ~ k, data=degree_df,log="xy")
lines(prob ~ k, data=theoretical_degree_df1)
lines(prob ~ k, data=theoretical_degree_df2)
```

In fact, people often fit stochastic block models (or more complicated variants)
to their data to detect communities. Because these models are generative, and have
a strong statistical foundation/justification, they are a nice choice for community
detection. There are a lot of R packages that will run an SBM for you, and most
do not come with extensive documentation - probably why you haven't used them before. 
Here's an example from one package that will do it for you (`randnet`).

```{r randnet}
pm <- cbind( c(.3, .01), c(.01, .5) )
sbm.net <- sample_sbm(n=100, pref.matrix=pm, block.sizes=c(30,70))

#Fit SBM with randnet package
#adapted from: https://methods.sagepub.com/dataset/howtoguide/sbm-in-karate-1977

# First get the optimal number of clusters
bhmc <- BHMC.estimate(as_adjacency_matrix(sbm.net),15)
bhmc

# Now fit your SBM with the desired number of clusters (really the "degree corrected SBM" according to their documentation)
sbm.fit <- reg.SSP(as_adjacency_matrix(sbm.net, sparse=FALSE), K=bhmc$K)
sbm.fit$cluster

plot(sbm.net, 
     vertex.label=NA,
     vertex.color = sbm.fit$cluster)

```

SBMs are in themselves a large field of research, so go do some reading.

## Bonus

Try looking at the degree distribution of the SpiecEasi network we made a couple
weeks ago (don't worry about edge weights). What shape does it have?